Real-space computation of $E$/$B$-mode maps I: Formalism, Compact Kernels, and Polarized Filaments

### 2018-07-20 articles/formalism/sections/discussion.tex
added references for r>0.001 and minor edits to text

These real space operators offer a spatially intuitive way of understanding the different decompositions of the Stokes vector on the sphere. We explicitly  demonstrated that all the real space operators are separable into azimuthal and radial parts. While the azimuthal part of the operators is primarily responsible for handling the spin decomposition, the radial weights determine the non-local dependence of the resulting fields on the original fields.  Only the radial part of the kernel depends on the band limit. 
The radial part of the operator kernel are roughly self-similar in the sense that the radial kernels evaluated with some band limit is related to another radial kernel evaluated with a different band limit by a simple rescaling of the function. We use this property to define a non-locality parameter $\beta_0$ as the angular distance at which the amplitude of the radial kernel falls below one percent of its maxima and it  is a function of the maximum multipole $\ell_{\rm max}$ available for the analysis. We empirically show that the non-locality parameter is approximately given by: $\beta_0 = \mathrm{min}(180^{\circ}, 180^{\circ} \frac{\ell_{0}}{\ell_{\rm max}})$ with $\ell_{0}=22$ for the operator that converts Stokes Q/U to scalars E/B and vice versa (see \fig{fig:rad_ker_decay}). \revisit{An analysis in  \cite{Zaldarriaga2001a} treated real space $E/B$ operators in the flat sky.  It did not explicitly derive the radial part of the kernel, but argued on geometric grounds that it should fall with angular separation as $\beta^{-2}$.  We find that this agrees with the average behavior of radial functions after averaging oscillations, and some such averaging takes place due to pixelization of the signal.  However, for precision reconstruction of $E/B$ on the sphere, the oscillations must be taken into account.}

Our careful study of the real space operators show that they can be expressed either as a Green's function or as a convolving beam function.  \revisit{The convolution interpretation is not a totally new concept.  It guides the discussion in \cite{Zaldarriaga2001a} and closely relates to the popular spokes and pinwheel descriptions of the $E/B$ modes.  However, the Green's function interpretation of the operators is a new one and is discussed here for the first time.} These two different interpretations of the operators emerge from the expression of the kernels in terms of the forward or inverse rotation Euler angles.

@@

%\revisit{Discussion on results from polarized filaments, compare and contrast with those in \cite{Zaldarriaga2001a}.}

@@

%Although the scaling for the real space method (${\cal O}(N_{\rm pix}^2)$) is steeper than for the spherical harmonic transforms (${\cal O}(N_{\rm pix}^{3/2})$), the simple parallelization and the small prefactor for compact kernels at high-resolution / high-band-limit could make it faster in some contexts.  It can have a small memory footprint.  \comment{Does parallel S2hat have memory constraints at high l?}  The evaluation of contribution from charges in the masked portion of the sky can simply be omitted.  \comment{hybrid low-l / high-l schemes}

\revisit{The real space kernels could in principle be incorporated into the pointing matrices for map making, allowing maps of $E/B$ to be made directly from instrument data, without the need for Stokes parameter maps as intermediate products.
  The pointing matrix---projecting the maps into the time-ordered data at a point---requires the convolution version of the kernel, while the transpose pointing matrix---projecting the time-ordered data to the map---requires the Green's function kernel.
  The method would be similar to pixel-based strategies to deconvolve an instrument beam during map making \citep[e.g.][]{2010ApJS..187..212C}.  However, such a strategy for map making would force the pointing matrix to be much less sparse and it may not be practical.}

@@

%\revisit{The discussion till now gives the impression that using the localized convolution kernels is no different from from using the default kernel and altering the spherical harmonic coefficients of expansion of the relavant fields by appropriately operating on them with the  effective beam functions $g_{\ell}$. To appreciate the difference between these two, it is important to realize that in general one can make a choice of a radial function which may not have a band limited description. In such a case these two method of evaluating the relevant fields is not identical. An example of this claim is depicted in \fig{fig:example_gbeta}.\\
%Another important thing to realize is that the harmonic coefficients derived from default full sky operations get some contributions from different portions of sky. For instance evaluating the E and B fields in the vicinity of the poles is are prone to receiving significant contributions from strong foregrounds near the equator. Correcting the harmonic coefficients of expansion with the effective beam function does not cancel these non-local contribution. On the contrary by performing the convolution with the localized real space kernels, the regions which contribute to the local field evaluations are predetermined by the choice of the radial function.}

>>>>

These real space operators offer a spatially intuitive way of understanding the different decompositions of the Stokes vector on the sphere. We explicitly  demonstrated that all the real space operators are separable into azimuthal and radial parts. While the azimuthal part of the operators is primarily responsible for handling the spin decomposition, the radial weights determine the non-local dependence of the resulting fields on the original fields.  Only the radial part of the kernel depends on the band limit. 
The radial part of the operator kernel are roughly self-similar in the sense that the radial kernels evaluated with some band limit is related to another radial kernel evaluated with a different band limit by a simple rescaling of the function. We use this property to define a non-locality parameter $\beta_0$ as the angular distance at which the amplitude of the radial kernel falls below one percent of its maxima and it  is a function of the maximum multipole $\ell_{\rm max}$ available for the analysis. We empirically show that the non-locality parameter is approximately given by: $\beta_0 = \mathrm{min}(180^{\circ}, 180^{\circ} \frac{\ell_{0}}{\ell_{\rm max}})$ with $\ell_{0}=22$ for the operator that converts Stokes Q/U to scalars E/B and vice versa (see \fig{fig:rad_ker_decay}). An analysis in  \cite{Zaldarriaga2001a} treated real space $E/B$ operators in the flat sky.  It did not explicitly derive the radial part of the kernel, but argued on geometric grounds that it should fall with angular separation as $\beta^{-2}$.  We find that this agrees with the average behavior of radial functions after averaging oscillations and some such averaging takes place due to pixelization of the signal.  However, for precision reconstruction of $E/B$ on the sphere, the oscillations must be taken into account.

Our careful study of the real space operators show that they can be expressed either as a Green's function or as a convolving beam function.  The convolution interpretation is not a totally new concept.  It guides the discussion in \cite{Zaldarriaga2001a} and closely relates to the popular spokes and pinwheel descriptions of the $E/B$ modes.  However, the Green's function interpretation of the operators is a new one and is discussed here for the first time. These two different interpretations of the operators emerge from the expression of the kernels in terms of the forward or inverse rotation Euler angles.

@@

@@

The real space kernels could in principle be incorporated into the pointing matrices for map making, allowing maps of $E/B$ to be made directly from instrument data, without the need for Stokes parameter maps as intermediate products.
  The pointing matrix---projecting the maps into the time-ordered data at a point---requires the convolution version of the kernel, while the transpose pointing matrix---projecting the time-ordered data to the map---requires the Green's function kernel.
  The method would be similar to pixel-based strategies to deconvolve an instrument beam during map making \citep[e.g.][]{2010ApJS..187..212C}.  However, such a strategy for map making would force the pointing matrix to be much less sparse and it may not be practical.

@@

### 2018-07-20 articles/formalism/sections/introduction.tex
added references for r>0.001 and minor edits to text

The primary aim of the current CMB experiments is to make precise measurements of the CMB polarizations. While the E-modes of polarization have been measured reasonably well by a number of experiments (\cite{2018RPPh...81d4901S} gives a recent review), the ongoing and future CMB experiments aim to measure the B-modes of CMB polarization with unprecedented accuracy and angular resolution. The detectors are swiftly approaching the desired sensitivities enabling us to in principle measure tiny $B$-mode signals ($r\gtrsim 0.001$) \cite{s4 science book, pico mission study}. However $B$-modes generated by galactic foreground are expected to be a few orders of magnitude higher than $B$-mode amplitude measurable by these detectors.  Precise modelling and subtraction of this large foreground contribution poses a major challenge for a robust unravelling of the minuscule $B$-mode signal.

>>>>

The primary aim of the current CMB experiments is to make precise measurements of the CMB polarizations. While the E-modes of polarization have been measured reasonably well by a number of experiments (\cite{2018RPPh...81d4901S} gives a recent review), the ongoing and future CMB experiments aim to measure the B-modes of CMB polarization with unprecedented accuracy and angular resolution. The instruments are swiftly approaching the desired sensitivities enabling us to in principle measure tiny $B$-mode signals ($r\gtrsim 0.001$) \cite{Spider, CLASS,Litebird, BICEP22015}. However $B$-modes generated by galactic foreground are expected to be a few orders of magnitude higher than $B$-mode amplitude measurable by these detectors.  Precise modelling and subtraction of this large foreground contribution poses a major challenge for a robust unravelling of the minuscule $B$-mode signal.

### 2018-07-20 articles/formalism/sections/kernel_discussion.tex
added references for r>0.001 and minor edits to text

These finding are neatly summarized in the observation that the radial functions with different maximum multipoles are approximately self-similar and related by this telescoping and scaling property:
$${}_rf(\beta,2,\ell_{\rm max}) \approx \Big[\frac{\ell_{\rm max}}{\ell'_{\rm max}}\Big]^2{}_rf(\beta'=\frac{\ell'_{\rm max}}{\ell_{\rm max}} \beta ,2,\ell'_{\rm max}) \qquad \revisit{\mbox{(small angles)}},$$
where ${}_rf , r \in [\mm, \md,\mi]$ denotes all the different radial functions. This telescoping property encapsulates both the amplitude scaling and leftward shift of the radial kernels on increasing the maximum multipole. \revisit{Specifically the global maxima of the kernel with maximum multipole $\ell_{max}$ is amplified(supressed) by a factor $(\ell_{\rm max}/\ell'_{\rm max})^2$ and the oscillating part is compressed(expanded) by a factor $(\ell'_{\rm max}/\ell_{\rm max})$ with respect to the kernel defined by the maximum multipole $\ell'_{\rm max}$ when $\ell_{\rm max} > \ell'_{\rm max}$($\ell_{\rm max} < \ell'_{\rm max}$).}

@@

The radial function ${}_\mm f(\beta)$ is oscillatory, regardless of the maximum $\ell_{\rm max}$. The envelope of the oscillation decays with angular separation at intermediate $\beta$ but increases as $\beta \rightarrow \pi$.  We find that the positive envelope of ${}_{\mm}f$ scales as $\beta^{-1.6}$ at intermediate $\beta$.  However, when we average over several oscillations with a moving window, we find that the average behavior at intermediate $\beta$ scales as $\beta^{-2}$, visible in \fig{fig:fbeta_avg}. We set the width of the smoothing window to $0.3\beta_0$, this results in the narrowing of the smoothing width as the band limit increases so as to keep the number of oscillations averaged over, remain roughly the same.  This agrees with the hypothesis for the \revisit{flat sky case} \comment{is this flat sky or continuum case?}, for which e.g. Zaldarriaga (2001) \cite{Zaldarriaga2001a} argued that form of the radial function has to be ${}_\mm f(\beta>0)\simeq\beta^{-2}$ to ensure that the Fourier modes of Stokes $Q/U$ relate to those of $E/B$ merely by rotations (with no scale dependence). In addition, we find that ${}_{\md}f$ under a moving average shows a similar scaling of $\beta^{-2}$, as seen in \fig{fig:dfbeta_avg}.
Applying a moving average on radial functions with increasing band limit, we see that a larger fraction of angular domain matches the scaling of $\beta^{-2}$.   Extrapolating this trend to the case of a very high band limit one expects this behavior to approach $\beta^{-2}$ across the domain $\beta\in[\epsilon,\pi-\epsilon]$ for $\epsilon \ll 1$. In essence the moving average can be understood as the averaging resulting from discrete sampling of a continuum field (one with an infinite band limit) where the width of the averaging window corresponds to the size of the pixel.

>>>>

These finding are neatly summarized in the observation that the radial functions with different maximum multipoles are approximately self-similar and related by this telescoping and scaling property:
$${}_rf(\beta,2,\ell_{\rm max}) \approx \Big[\frac{\ell_{\rm max}}{\ell'_{\rm max}}\Big]^2{}_rf(\beta'=\frac{\ell'_{\rm max}}{\ell_{\rm max}} \beta ,2,\ell'_{\rm max}) \qquad ,$$
where ${}_rf , r \in [\mm, \md,\mi]$ denotes all the different radial functions. This telescoping property encapsulates both the amplitude scaling and leftward shift of the radial kernels on increasing the maximum multipole. \revisit{Specifically the global maxima of the kernel with maximum multipole $\ell_{max}$ is amplified by a factor $(\ell_{\rm max}/\ell'_{\rm max})^2$ and the oscillating part is compressed by a factor $(\ell'_{\rm max}/\ell_{\rm max})$ with respect to the kernel defined by the maximum multipole $\ell'_{\rm max}$ when $\ell_{\rm max} > \ell'_{\rm max}$. The details of the accuracy of this approximation vary with the angular separation $\beta$ and the difference in the maximum multipoles ($\ell_{\rm max}$ \& $\ell'_{\rm max}$)  used in evaluating the two functions.}

@@

The radial function ${}_\mm f(\beta)$ is oscillatory, regardless of the maximum $\ell_{\rm max}$. The envelope of the oscillation decays with angular separation at intermediate $\beta$ but increases as $\beta \rightarrow \pi$.  We find that the positive envelope of ${}_{\mm}f$ scales as $\beta^{-1.6}$ at intermediate $\beta$.  However, when we average over several oscillations with a moving window, we find that the average behavior at intermediate $\beta$ scales as $\beta^{-2}$, visible in \fig{fig:fbeta_avg}. We set the width of the smoothing window to $0.3\beta_0$, this results in the narrowing of the smoothing width as the band limit increases so as to keep the number of oscillations averaged over, remain roughly the same.  This agrees with the hypothesis for the \revisit{flat sky and continuum case}, for which e.g. Zaldarriaga (2001) \cite{Zaldarriaga2001a} argued that form of the radial function has to be ${}_\mm f(\beta>0)\simeq\beta^{-2}$ to ensure that the Fourier modes of Stokes $Q/U$ relate to those of $E/B$ merely by rotations (with no scale dependence). In addition, we find that ${}_{\md}f$ under a moving average shows a similar scaling of $\beta^{-2}$, as seen in \fig{fig:dfbeta_avg}.
Applying a moving average on radial functions with increasing band limit, we see that a larger fraction of angular domain matches the scaling of $\beta^{-2}$.   Extrapolating this trend to the case of a very high band limit one expects this behavior to approach $\beta^{-2}$ across the domain $\beta\in[\epsilon,\pi-\epsilon]$ for $\epsilon \ll 1$. In essence the moving average can be understood as the averaging resulting from discrete sampling of a continuum field (one with an infinite band limit) where the width of the averaging window corresponds to the size of the pixel.
